how to \texttt{git bisect} when your test suite is having a bad day

### 2020-03-29 bisect/paper/paper.tex
write more

Let's say you're trying to bisect
% a bug in
a given range of $n$ commits.%
\footnote{To use binary search to find a commit that introduced a bug.}
Call them $c_0 \dots c_{n-1}$, where $c_0$ is known safe and $c_n$ is known to have the bug.
You'd probably start by testing $c_{n/2}$, right?
And you'd expect to pinpoint the buggy comit in $\log_2(n)$ steps.
That's math.

@@

Ok, so let's model the problem as a sequence of steps on a probability distribution function
(henceforth, PDF; and also, CDF for the cumulative kind).
Initially, $\pdf(i) = 1/n$ for all $0 \le i < n$.
When the bug reproduces at some commit $b$,
you're certain no $c_{i>b}$ introduced the bug, so $\pdf(i>b) = 0$,
and $\pdf(i\le b)$ renormalizes as $1/b$, to satisfy the $\int \pdf(i) \text{d} i = 1$ invariant.%
\footnote{Implemented as {\tt fn adjust\_pdf\_bug\_repros()} in the simulator.}

In the deterministic case ($p=1$),
the symmetric thing happens when the test passes at some $c_j$: $\pdf(i\le j) = 0$.
But when $p<1$, we must generalized it \cite{mario3}. Bayes's rule provides that:
\[
P(A|B) = \frac{P(B|A) \times P(A)}{P(B)}
\]

@@

In this case, $B$ is that the test passed at $c_j$, and $A$ is that bug be present at or before $c_j$ after all.
$P(B|A)$ is simply $1-p$.
$P(A)$ is the prior on $c_j$ containing the bug, i.e., $\cdf(j)$.
And $P(B)$ is the false negative probability weighted by the bug existing, i.e., $(1-p)\cdf(j) + 1(1-\cdf(j))$.
% simplifies to: cdf(j) - p*cdf(j) + 1 - cdf(j)
% simplifies to: 1 - p*cdf(j)
To update our priors on commits up to $j$, we renormalize by dividing out the old $\cdf(j)$
and multiplying by the new $P(A|B)$, i.e.,
%To update our priors on commits up to $j$, we divide out the old $\cdf(j)$
%and multiply by the new $P(A|B)$.
%Call this renormalization factor $\renorm{j}{i}$.
%\[
%	\forall i \le j \Rightarrow \renorm{i}{i} = 
%	\frac{1}{\cdf(j)}
%	\frac{(1-p)\cdf(j)}{(1-p)\cdf(j) + (1-\cdf(j))}
%\]
\[
	\forall i \le j, \pdf'(i)
	\leftarrow
	\pdf(i)
	\frac{1}{\cdf(j)}
	\frac{(1-p)\cdf(j)}{(1-p)\cdf(j) + (1-\cdf(j))}
\]
Which simplifies to:
\[
	\forall i \le j, \pdf'(i)
	\leftarrow
	\pdf(i)
	\frac{1-p}{1 - p\cdf(j)}
\]
Call this renormalization factor $\renorm{}$.
As a sanity check, $p\cdf(j)$ is less than $p$, so $\renorm{i\le j} < 1$.

@@

Now let's say the test passes at $j=9$, then at $j=10$.
\Cref{fig:example}(b) shows the updated PDFs/CDFs:
for $p=1$, this pinpoints the bug at $j=11$, and the search is over.
But for $p=0.5$, there's still 2/3 odds we'd be wrong!
In fact, from here it takes {\it eighteen} further probes at $j=10$ until we are at least five 9's confident that $c_{11}$ is the culprit.%
\footnote{See {\tt fn test\_figure\_1()}!}

@@

Dropbox has a CI service called Athena \cite{athena}
which automatically searches for flaky tests, when otherwise no bug exists.
In this case the goal is to keep the build green, but if you consider the flaky test itself to be the bug, it's the same problem.%
\footnote{Incidentally, the symmetric case
-- where a bug repros with $p=1$, but the test also flakes with some $q<1$ --
is also the same problem.}
Athena's ``deflakes'' the test at each commit by running 10 times,
and treats the combined result as ``basically as good as $p=1$''.%
\footnote{Upcoming, I will call this strategy $\mathsf{Mistrustful(10)}$.}
In this setting, $p$ is not known in advance, so using Bayes's rule is off the table.
But I will show that even without access to the PDF, a better strategy exists. % at least by my definition

@@

Ok, so how do we
%cope with the fact that we'll never have a proper lower bound?
%How do we
make progress,
i.e., concentrate probability mass til there's $\confidents$ of it on one commit,
as quickly as possible?
%In principle, testing anywhere in the range $[0,b)$,
%where $b$ is the earliest known buggy commit,
%will make progress,
%
%% ghhh this sentence only feels good if i know what i'm talking about
%Let's deconstruct the motives of classical binary search.
Let $c_b$ be the earliest known buggy commit, and $c_a$ be the latest known safe commit.
In Determinism World, we would bisect at $c_{(a+b)/2}$
because that minimizes the maximum range remaining in the worst case outcome
(i.e., the two possible outcome ranges are the same length).
%But in Nondeterminism World,
%the worst case is a big constant times linear:
%an adversarial RNG god could pass your test a thousand times on $c_{n-1}$, then fail,
%then pass it a thousand times on $c_{n-2}$, then fail, and so on,
%before ultimately arriving at buggy commit $c_0$.
%So thinking about the worst case is not productive for a sublinear strategy,
%and also,
%we shouldn't be thinking about the range length,
%because a pass is still progress even though $a$ doesn't ever budge,
But in Nondeterminism World, $a$ doesn't ever budge from $0$,
so bisecting at $c_{(0+b)/2}$ will not even terminate.
%will eventually fail to make progress.
Sure, hitting the PDF with
%Bayes's rule at $b/2$
$\renorm{b/2}$
will always move {\it some} mass rightward,
but once five 9's of it is already over there,
%to the right of that point,
it can't concentrate it onto one point.
So let's not think about the range.

@@

Another way to think about Determinism World is the $c_{(a+b)/2}$ bisect point
% maximizes the expected probability mass that moves from one side to the other.
cuts the probability mass, rather than the range, in half,
% How about in Nondeterminism World,
% we ignore the range length, and instead bisect at the midpoint of probability mass,
i.e., $\mathsf{last}_i \cdf(i) \le 0.5$.
This approach is compatible with Nondeterminism World,
fits the ``binary search'' spirit,
and will at least terminate
(if $b$ is the solution, it converges to repeatedly probing $b-1$),
which is more than you can say for the range-based $c_{(0+b)/2}$.
%It is intuitively obvious that this will make progress faster than bisecting at $c_{(b+0)/2}$,
%But is it fast?
However, it doesn't maximize either the minimum or the expected amount of mass that moves:
when $p<1$, it's intuitively easy to see that CDF threshold should actually be less than $0.5$.
\footnote{I did some pen-and-paper algebra to find a closed form for this,
but I actually trust the simulator more than myself at this point,
so I stopped spending brains on this part of the problem. I kind of regret that.}
% TODO TODO: actually do the paperwork.
Is maximizing the expected moved mass as a function of $p$ the fastest we can go,
and if so, why?

@@

A PDF's information content is measured in {\it entropy}:
$\entropy = \sum_i \pdf(i)\mathsf{ln}(\pdf(i))$.
Indeed,
in Determinism World the search terminates when $\entropy = 0$.
I thought for a while about how to link ``minimum expected entropy''
to the stated goal %in Nondeterminism World:
of
%minimizing the expected steps until
$\confidents = \mathsf{max}_i \pdf(i) \ge 0.99999$,
but couldn't really formalize the idea.
The goal of five 9's is fairly arbitrary anyway,
and also not necessarily stable under the entropy measurement,
since it doesn't care how the remaining $0.00001$ is distributed.
%
Also, a Very Smart Math Friend of mine was unable to find a simple closed form for
the expected resulting $\entropy(\renorm{} \circ \pdf)$
%\[
%(\cdf(j) + (1-\cdf(j))(1-p))\entropy(\renorm \circ \pdf)
%+
%p(1-\cdf(j))\entropy(\renorm{\text{bug}} \circ \pdf)
%\]
or whatever (bleh, my notation is kinda reaching its limits here).
%so how would you write a proof about it anyway.
Nevertheless, it seems promising, and easy to implement.

@@

You can tell it's at this point I'm starting to get statistics fatigue.

@@

@@

%(
% hmmm...

% In Determinism World, the search terminates when

% Let's think of bisecting at $c_{(a+b)/2}$, in Determinism World, 
% as maximizing the expected information gained from either test result.
% On PDFs, information gain is measured as

@@

\cite{drowningpoint}
\cite{epsilon}

@@

% weigh cost of running test suite vs cost of git checkout -- mistrustful() gets a whole lot better! but, where is the break even point when you should git checkout?

% p not known, figur eit out on the fly

@@

>>>>

Let's say you're trying to bisect
% a bug in
a given range of $n$ commits.%
\footnote{To use binary search to find a commit that introduced a bug.}
Call them $c_0 \dots c_{n-1}$, where $c_0$ is known safe and $c_n$ is known to have the bug.
You'd probably start by testing $c_{n/2}$, right?
And you'd expect to pinpoint the buggy comit in $\mathsf{log} (n)$ steps.
That's math.

@@

Ok, so let's model the problem as a sequence of steps on a probability distribution function
(henceforth, PDF; and also, CDF for the cumulative kind).
Initially, $\pdf(i) = 1/n$ for all $0 \le i < n$.
When the bug reproduces at some commit $b$,
you're certain no $c_{i>b}$ introduced the bug, so each $\pdf(i>b)$ becomes $0$,
and each $\pdf(i\le b)$ renormalizes by a factor of $n/b$, to preserve the $\int \pdf(i) \text{d} i = 1$ invariant.%
\footnote{Implemented as {\tt fn adjust\_pdf\_bug\_repros()} in the simulator.}

In the deterministic case ($p=1$),
the symmetric thing happens when the test passes at some $c_j$: $\pdf(i\le j) = 0$.
But when $p<1$, we must generalized it \cite{mario3}. Here's Bayes's rule:
\[
P(A|B) = \frac{P(B|A) \times P(A)}{P(B)}
\]

@@

In this case, $B$ is that the test passes at $c_j$, and $A$ is that bug exists at or before $c_j$ after all.
$P(B|A)$ is the false negative rate, i.e. $1-p$.
$P(A)$ is the prior on $c_j$ containing the bug, i.e., $\cdf(j)$.
And $P(B)$ is the false negative rate weighted by the bug existing, i.e., $(1-p)\cdf(j) + 1(1-\cdf(j))$.
% simplifies to: cdf(j) - p*cdf(j) + 1 - cdf(j)
% simplifies to: 1 - p*cdf(j)
To update our priors on commits up to $j$, we renormalize by dividing out the old $\cdf(j)$
and multiplying by the new $P(A|B)$, i.e.,
%To update our priors on commits up to $j$, we divide out the old $\cdf(j)$
%and multiply by the new $P(A|B)$.
%Call this renormalization factor $\renorm{j}{i}$.
%\[
%	\forall i \le j \Rightarrow \renorm{i}{i} = 
%	\frac{1}{\cdf(j)}
%	\frac{(1-p)\cdf(j)}{(1-p)\cdf(j) + (1-\cdf(j))}
%\]
\[
	\forall i \le j, \pdf'(i)
	\leftarrow
	\pdf(i)
	\frac{1}{\cdf(j)}
	\frac{(1-p)\cdf(j)}{(1-p)\cdf(j) + (1-\cdf(j))}
\]
Which simplifies to:
\[
	\forall i \le j, \pdf'(i)
	\leftarrow
	\pdf(i)
	\frac{1-p}{1 - p\cdf(j)}
\]
Call this renormalization factor $\renorm{}$.
As a sanity check, $p\cdf(j)$ is less than $p$, so $\renorm{i\le j} < 1$.

@@

Now let's say the test passes at $j=9$, then at $j=10$.
\Cref{fig:example}(b) shows the updated PDFs/CDFs:
for $p=1$, this pinpoints the bug at $j=11$, and the search is over.
But for $p=0.5$, there's still 2/3 odds we'd be wrong!
In fact, from here it takes 18 further probes at $j=10$ until we are at least five 9's confident that $c_{11}$ is the culprit.%
\footnote{See {\tt fn test\_figure\_1()}!}
Bayes's rule gonna get ya.

@@

Dropbox has a CI service called Athena \cite{athena}
which automatically searches for flaky tests, when otherwise no bug exists.
In this case the goal is to keep the build green, but if you consider the flaky test itself to be the bug, it's the same problem.%
\footnote{Incidentally, the symmetric case
-- where a bug repros with $p=1$, but the test also flakes with some $q<1$ --
is also the same problem.}
Athena's ``deflakes'' the test at each commit by running 10 times,
treating the combined result as ``basically as good as $p=1$'',
and then runs an otherwise classical binary search.%
\footnote{Upcoming, I will call this strategy \textsf{Human-Mistrustful}$(10)$.}
In this setting, $p$ is not known in advance, so using Bayes's rule is off the table.
But I will show that even without access to the PDF, a better strategy exists. % at least by my definition

@@

Ok, so how do we
%cope with the fact that we'll never have a proper lower bound?
%How do we
make progress,
i.e., concentrate probability mass til there's $\confidents$ of it in one place,
as quickly as possible?
%In principle, testing anywhere in the range $[0,b)$,
%where $b$ is the earliest known buggy commit,
%will make progress,
%
%% ghhh this sentence only feels good if i know what i'm talking about
Let's deconstruct the motives of classical binary search.
Let $c_b$ be the earliest known buggy commit, and $c_a$ be the latest known safe commit.
In Determinism World, bisecting at $c_{(a+b)/2}$
minimizes the worst case maximum range remaining,
as the two possible outcome ranges are the same length.
%But in Nondeterminism World,
%the worst case is a big constant times linear:
%an adversarial RNG god could pass your test a thousand times on $c_{n-1}$, then fail,
%then pass it a thousand times on $c_{n-2}$, then fail, and so on,
%before ultimately arriving at buggy commit $c_0$.
%So thinking about the worst case is not productive for a sublinear strategy,
%and also,
%we shouldn't be thinking about the range length,
%because a pass is still progress even though $a$ doesn't ever budge,
But in Nondeterminism World, $a$ doesn't ever budge from $0$,
so bisecting at $c_{(0+b)/2}$ will not even terminate.
%will eventually fail to make progress.
Sure, hitting the PDF with
%Bayes's rule at $b/2$
$\renorm{b/2}$
will always move {\it some} mass rightward,
but once five 9's of it is already over there,
%to the right of that point,
it can't concentrate it onto one point.
So let's not think about the range.

@@

Another way to think about Determinism World is that the $c_{(a+b)/2}$ bisect point
% maximizes the expected probability mass that moves from one side to the other.
cuts the probability mass, rather than the range, in half,
% How about in Nondeterminism World,
% we ignore the range length, and instead bisect at the midpoint of probability mass,
i.e., $\mathsf{max}_j (i)$ s.t. $\cdf(j) \le 0.5)$.
This approach
fits the ``binary search'' spirit,
and will terminate in Nondeterminism World,
i.e.,
if $b$ is the solution, it converges to repeatedly probing $b-1$,
so it can pass any fixed $\confidents$ threshold.
%(which is more than you can say for the range-based $c_{(0+b)/2}$).
%It is intuitively obvious that this will make progress faster than bisecting at $c_{(b+0)/2}$,
%But is it fast?
% However, it doesn't maximize either the minimum or the expected amount of mass that moves:
% when $p<1$, it's intuitively apparent that this CDF threshold should actually be less than $0.5$.%
%\footnote{I did some pen-and-paper algebra to find a closed form for this,
%but I actually trust the simulator more than myself at this point,
%so I stopped spending brains on this part of the problem. I kind of regret that.}
%
%Is maximizing the expected moved mass as a function of $p$ the fastest we can go,
%and if so, why?
But is $0.5$ still the best bisect point even when $p<1$?
I wondered if this could be expressed as the amount of probability mass moved from one side to the other,
i.e., $\sum_i \mathsf{abs}(\pdf(i) - \pdf'(i))$.
In the case where the bug repros this is:
\[
	p\cdf(j) \times (1-\cdf(j))
\]
and in the case where the test passes:
\[
	(1-p\cdf(j)) \times \cdf(j) \times (1-\renorm{i\le j})
\]
which, surprisingly, simplifies to exactly the same thing as the bug repros case.
The maximum occurs where $\partial/\partial \cdf(j)$ is 0,
which turns out to be at $\cdf(j)=0.5$ after all, and independent of $p$.%
\footnote{I also worked out the {\it minimum} moved mass between the pass and fail case,
which is almost always the pass case,
and comes out to
%$\frac{p\cdf(j)(1-\cdf(j))}{1-p\cdf()}$
$\frac{\cdf(j)(1-\cdf(j))}{1/p-\cdf(j)}$,
which experiences its maximum at
$\frac{1-\sqrt{1-p}}{p}$ (thanks wolframalpha).
But this is worst-case thinking, which doesn't seem appropriate.
Let's have some optimism!}
But it's not clear that the amount of mass moved
necessarily corresponds to reaching $\confidents$ the fastest.
%It just sounds kinda nice?
%What is the best threshold, perhaps as a function of $p$, and why?

%expected moved entropy
%	call cdfj 'z' for the threshold we're looking for
%
%moved entropy at j in pass case
%	new area is z (1-p)/(1-pz)
%	old area is z
%	so:
%	z - z R
%	z (1-R)
%
%	z [ 1-pz - (1-p) ] / (1-pz)
%	z [ p-pz ] / (1-pz)
% moved entropy to the right of j, should be equal
% 	new area is (1-z)R
% 		is (1-z)/(1-pz)
% 	old area is 1-z
% 	moved area is (1-z)(R-1)
% 		is (1-z)(pz)/(1-pz)
% 		is (pz-pzz)/(1-pz) yup, checks out
% 
%
%probability that it moves = p test passes
%	1(1-z) + (1-p)z (that's B)
%	= 1-z + z-pz
%	= 1-pz (still B)
%
%so this term:
%	(1-pz) (z) (p-pz) / (1-pz)
%	= z(p-pz)
%	= pz-pzz
%
%moved entropy at j in fail case
%	new area is 1
%	old area is z
%	of course, all 1-z of that shit moves
%
%probability that it moves = p test fails
%	pz
%
%so this term:
%	pz(1-z)
%	= pz-pzz
%	(wow, that's the same?)
%
%total expected moved: 2pz-2pzz
%
%solve for derivative dz = 0 to find argmax z
%	p - 2pz = 0

@@

A PDF's information content is measured in entropy:
$\entropy = \sum_i \pdf(i)\mathsf{ln}(\pdf(i))$.
Alone, this value is fairly meaningless, but to compare them,
a spikier PDF has lower entropy than a flatter one.
Indeed,
when the search terminates in Determinism World, $\entropy = 0$.
I thought for a while about how to link ``minimum expected entropy''
to the stated goal %in Nondeterminism World:
of
%minimizing the expected steps until
$\confidents = \mathsf{max}_i (\pdf(i)) \ge 0.99999$,
but couldn't really formalize the idea.
The goal of five 9's is fairly arbitrary anyway,
and also not necessarily stable under the entropy measurement,
since it doesn't care how the remaining $0.00001$ is distributed.%
\footnote{See {\tt fn test\_entropy\_stability()}.}

This strategy is expensive to compute.
Whereas bisecting at some fixed threshold of probability mass merely requires a $O(\mathsf{log}  n)$ query of the CDF,
computing the expected entropy is $O(n)$ for {\it each} bisect point.
A Very Smart Math Friend of mine \cite{primer} analyzed the easy case of the initial test,
when the PDF is still uniform (i.e., $\forall i, \cdf(i) = i$),
and found the closed form
\[
	\entropy_{j,0} = p \mathsf{ln}(j)
	+
	(1-pj) \mathsf{ln} (1-pj)
	-
	j(1-p) \mathsf{ln} (j-p)
\]
whose derivative,
in his words,
``looks like a giant mess that does not admit an analytic solution for $j$.''

% even the easy case of the inn
% %the expected resulting
% $\mathsf{EV}(\entropy(\renorm{} \circ \pdf))$
% %\[
% %(\cdf(j) + (1-\cdf(j))(1-p))\entropy(\renorm \circ \pdf)
% %+
% %p(1-\cdf(j))\entropy(\renorm{\text{bug}} \circ \pdf)
% %\]
% or whatever (bleh, my notation is kinda reaching its limits here).
% %so how would you write a proof about it anyway.
% Nevertheless, it seems promising, and easy to implement.

There is a ray of hope, however:
thanks to the nondecreasing-PDF invariant I mentioned on the previous page,
the expected entropy has a unique local minimum.%
\footnote{See {\tt fn test\_expected\_entropy\_has\_unique\_local\_minimum()}.}
Thus we can binary search for the global minimum, making this at least $O(n\mathsf{log}n)$
instead of $O(n^2)$.

@@

You can tell at this point I'm starting to get statistics fatigue.

@@

% parameterize over 10
% what should the human do when they think they're done? -> probe repeatedly at b-1
% what should they do when

@@

@@

To save you some page-turning, here's the source code link again:
\url{https://github.com/bblum/sigbovik/blob/master/bisect}.
A {\tt BisectStrategy}
implements a function over $p$, the PDF, and/or its own internal state
to choose the next bisect point.
The {\tt SimulationState} is initialized with fixed $n$ and $p$,
and repeatedly invokes a given {\tt BisectStrategy},
hitting the PDF with Bayes's rule accordingly,
until $\confidents \ge 0.99999$.
I provide implementations for all the strategies from the previous section.
It's written in Rust so it's self-documenting.

I learned a lot about floating point.
I don't mean like NaNs and mantissa bits,
I mean that when your problem is fundamentally a long multiplication sequence of increasingly high rationals,
your precious $\cdf(n)$ inevitably drifts ever farther from 1, and you start drowning in imprecision \cite{drowningpoint}.
I suppose I could have used the {\tt BigRational} library,
but I chose Rust over Haskell for this so I could {\it avoid} runaway memory consumption,
so I resigned myself to explicitly renormalizing my PDFs by $1/\cdf(n)$ each time I updated them.
After that, I was able to write some assertions
to bound the imprecision drift within $k(\text{\tt std::f64::EPSILON})$ \cite{epsilon}.
But yeah, I definitely spent some sanity points writing a function named {\tt fn assert\_kinda\_equals\_one(\&self)}.

@@

I'm not ashamed to admit I left a lot of loose ends in this paper.
The biggest open questions for me are:
\begin{enumerate}
	\item What relates $p$ to {\sf cdfbisect}'s optimal argument?
	\item Why is {\sf entropy} {\it so tantalizingly close} to {\sf cdfbisect}? Is there some way to show they're the same?
\end{enumerate}
I have also prepared a set of backup questions in case those aren't enough for you:
\begin{enumerate}
	\setcounter{enumi}{3}
	\item If you don't know $p$ in advance, is there an efficient way to multiplex the search with figuring it out on the fly?
	\item What if false positives are also possible, with some $q \ne p$?
	\item How do you account for the cost of {\tt git checkout} being nonzero (i.e., having to recompile your tests on new code)?
		Clearly {\sf human-mistrustful} becomes better. Where is the break-even point?
\end{enumerate}

@@

If you are a computer, you can do pretty well by bisecting the probability mass somewhere between 1/3 and 1/2.
%
If you are a human, you should forget everything you know as soon as you see new evidence that contradicts your priors.

And remember to be patient and kind. Your test suite is just trying its best!
